---
title: "Pipes with dplyr"
subtitle: "Stat 133, Spring 2018"
author: "Gaston Sanchez"
output: github_document
fontsize: 11pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, fig.path = '05-images/')
library(knitr)
library(dplyr)
library(ggplot2)
library(magrittr)
```

> ### Learning Objectives:
>
> - Compare base R and `"dplyr"`
> - Get to know the pipe operator `%>%`

------

## Introduction

Last week you started to manipulate data tables (e.g. `data.frame`, `tibble`)
with functions provided by the R package `"dplyr"`.

Having been exposed to the _dplyr_ paradigm, let's compare R base manipulation against the various dplyr syntax flavors.


### Starwars Data Set

In this tutorial we are going to use the data set `starwars` that comes in `"dplyr"`:

```{r warning = FALSE, message = FALSE}
# load dplyr
library(dplyr)

# data set
starwars
```


### Average Height of Male and Female Individuals

For illustration purposes, let's consider a relatively simple example.
Say we are interested in calculating the average (mean) height for both female 
and male individuals. Let's discuss how to find the solution under the base R
approach, as well as the dplyr approach.

-----


## Quick inspection of `height`

```{r}
# summary stats of height
summary(starwars$height)
```

```{r height_histogram}
# histogram
hist(starwars$height, col = 'gray80', las = 1)
```


### Quick inspection of `gender`

```{r}
# frequencies of gender
summary(starwars$gender)
gender_freqs <- table(starwars$gender)
gender_freqs
```

```{r gender_barchart}
# barchart of gender freqs
barplot(gender_freqs, border = NA, las = 1)
```

Now let's use `"dplyr"` to get the frequencies:

```{r}
# distinct values 
distinct(starwars, gender)
```

Oh! Notice that we have some missing values, which were not reported by `table()`.

```{r}
# frequencies of gender (via dplyr) 
count(starwars, gender)
```


-----


## Base R approach

Let's see how to use base R operations to find the average `height` of individuals with `gender` female and male.

```{r}
# identify female and male individuals
# (comparison operations)
which_females <- starwars$gender == 'female'
which_males <- starwars$gender == 'male'
```

```{r}
# select the height values of females and males
# (via logical subsetting)
height_females <- starwars$height[which_females]
height_males <- starwars$height[which_males]
```

```{r}
# calculate averages (removing missing values)
avg_ht_female <- mean(height_females, na.rm = TRUE)
avg_ht_male <- mean(height_males, na.rm = TRUE)

# optional: display averages in a vector
c('female' = avg_ht_female, 'male' = avg_ht_male)
```


All the previous code can be written with more compact expressions:

```{r}
# all calculations in a couple of lines of code
c("female" = mean(starwars$height[starwars$gender == 'female'], na.rm = TRUE),
  "male" = mean(starwars$height[starwars$gender == 'male'], na.rm = TRUE)
)
```


-----


## With `"dplyr"`

The behavior of `"dplyr"` is functional in the sense that function calls don't 
have side-effects. You must always save their results in order to keep them 
in an object (in memory). This doesn't lead to particularly elegant code, 
especially if you want to do many operations at once.


### Option 1) Step-by-step

You either have to do it step-by-step:

```{r}
# manipulation step-by-step
gender_height <- select(starwars, gender, height)

fem_male_height <- filter(gender_height, 
                          gender == 'female' | gender == 'male')

height_by_gender <- group_by(fem_male_height, gender)

summarise(height_by_gender, mean(height, na.rm = TRUE))
```


### Option 2) Nested (embedded) code

Or if you don't want to name the intermediate results, you need to wrap the 
function calls inside each other:

```{r}
summarise(
  group_by(
    filter(select(starwars, gender, height),
           gender == 'female' | gender  == 'male'),
    gender),
  mean(height, na.rm = TRUE)
)
```

This is difficult to read because the order of the operations is from inside 
to out. Thus, the arguments are a long way away from the function. 


### Option 3) Piping

To get around the problem of nesting functions, `"dplyr"` also provides the 
`%>%` operator from the R package `"magrittr"`.

What does the _piper_ `%>%` do? Here's a conceptual example: 

```{r eval = FALSE}
x %>% f(y)
```

`x %>% f(y)` turns into `f(x, y)` so you can use it to rewrite multiple 
operations that you can read left-to-right, top-to-bottom.

Here's how to use the piper to calculate the average height for female and 
male individuals:

```{r}
avg_height_by_gender <- starwars %>% 
  select(gender, height) %>%
  filter(gender == 'female' | gender == 'male') %>%
  group_by(gender) %>%
  summarise(avg = mean(height, na.rm = TRUE))

avg_height_by_gender

avg_height_by_gender$avg
```

-----

## Another Example

Here's another example in which we calculate the mean `height` and mean `mass` of `species` Droid, Ewok, and Human; arranging the rows of the tibble by mean height, in descending order:

```{r}
starwars %>%
  select(species, height, mass) %>%
  filter(species %in% c('Droid', 'Ewok', 'Human')) %>%
  group_by(species) %>%
  summarise(
    mean_height = mean(height, na.rm = TRUE),
    mean_mass = mean(mass, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_height))
```

-----

## Pipes and Plots

You can also the `%>%` operator to chain dplyr commands with ggplot commans (and other R commands). The following examples combine some data manipulation to `filter()` female and males individuals, in order to graph a density plot of `height`

```{r densities}
starwars %>%
  filter(gender %in% c('female', 'male')) %>%
  ggplot(aes(x = height, fill = gender)) + 
  geom_density(alpha = 0.7)
```

Here's another example in which instead of graphing density plots, we graph boxplots of `height` for female and male individuals:

```{r boxplots}
starwars %>%
  filter(gender %in% c('female', 'male')) %>%
  ggplot(aes(x = gender, y = height, fill = gender)) + 
  geom_boxplot()
```

-----

## More Pipes

Often, you will work with functions that don't take data frames (or tibbles) as
inputs. A typical example is the base `plot()` function used to produce a 
scatterplot; you need to pass vectors to `plot()`, not data frames. In this 
situations you might find the `%$%` operator extremely useful.

```{r eval = FALSE}
library(magrittr)
```

The `%$%` operator, also from the package `"magrittr"`, is a cousin of the 
`%>%` operator. What `%$%` does is to _extract_ variables in a data frame
so that you can refer to them explicitly. Let's see a quick example:

```{r scatterplot}
starwars %>%
  filter(gender %in% c('female', 'male')) %$%
  plot(x = height, y = mass, col = factor(gender), las = 1)
```
