---
title: "Getting started with functions"
subtitle: "Stat 133, Spring 2018"
author: "Gaston Sanchez"
output: github_document
fontsize: 11pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```


> ### Learning Objectives
>
> - Define a function that takes arguments
> - Return a value from a function
> - Test a function
> - Set default values for function arguments
> - Documenting a function

------

## Motivation

- R comes with many functions (and packages) that let us perform a wide variety of tasks.
- Most of the things we do in R is via calling some function.
- Sometimes, however, there's no function to do what we want to achieve.
- When that's the case, you will want to write your own functions.

So far you've been using a number of functions in R. Now it's time to see 
how you can create and use your own functions.

Consider the data set `starwars` that comes in the package `"dplyr"`

```{r}
starwars
```

Let's focus on the variable `height`, more specifically on the first 10 values:

```{r}
ht10 <- starwars$height[1:10]
ht10
```

The values of `height` (and `ht10`) are expressed in centimeters, but what if we wanted to obtain values in inches? The conversion formula is 1 cm = 0.3937 in.

```{r}
# height in inches
ht10 * 0.3937
```

This works. But what if you had more data sets, all of them containing `height` values in cms, and you needed to convert those cms into inches? Wouldn't be nice to have a dedicated function `cm2in()`?

```r
cm2in(ht10)
```

R does not have a built-in function `cm2in()` but we can create one. Let's see how to do it "logically" step by step:

```{r results='hide'}
# 1) concrete example
ht10 * 0.3937

# 2) make it more general
x <- ht10
y <- x * 0.3937

# 3) encapsulate code with an R expression
{
  y <- x * 0.3937
}

# 4) create function
cm2in <- function(x) {
  y <- x * 0.3937
  return(y)
}

# 5) test it
cm2in(ht10)

# 6) keep testing
cm2in(starwars$height)
```

- To define a new function in R you use the function `function()`.
- You need to specify a name for the function, and then assign `function()`
to the chosen name. 
- You also need to define optional arguments (i.e. inputs). 
- And of course, you must write the code (i.e. the body) so the function does 
something when you use it:


-----

## Anatomy of a function

```{r}
# anatomy of a function
some_name <- function(arguments) {
  # body of the function
}
```

- Generally, you give a name to a function.
- A function takes one or more inputs (or none), known as _arguments_.
- The expressions forming the operations comprise the __body__ of the function.
- Usually, you wrap the body of the functions with curly braces.
- A function returns a single value.

A less abstract function could have the following structure:

```r
some_name <- function(arg1, arg2, etc) 
{
  expression_1
  expression_2
  ...
  expression_n
}
```

-----

### Scale Transformations

Let's see another example. Often, we need to transform the scale of one or more variables. Perhaps the most common type of transformation is when we _standardize_ a variable, that is: mean-center and divide by its standard deviation:

![standard score](https://wikimedia.org/api/rest_v1/media/math/render/svg/5ceed701c4042bb34618535c9a902ca1a937a351)

R has the function `scale()` that can be used to perform this operation, but let's pretend for a minute that there's no function in R to calculate standard scores. Here are the primary steps to compute such score:

- compute the mean
- compute the standard deviation
- calculate deviations from mean
- divide by standard deviation

```{r}
x <- ht10
x_mean <- mean(x)
x_sd <- sd(x)
x_centered <- x - x_mean
z <- x_centered / x_sd
z
```

Having the code of the body, we can encapsulate it with a function:

```{r}
# first round
standardize <- function(x) {
  x_mean <- mean(x)
  x_sd <- sd(x)
  x_centered <- x - x_mean
  z <- x_centered / x_sd
  return(z)
}
```

And now we can test it:

```{r}
standardize(ht10)
```

What about applying `standardize()` on the entire column `height`:

```{r}
standardize(starwars$height)
```

Ooops! Because `starwars$height` contains missing values, our `standardize()` function does not know how to deal with them.


### Dealing with missing values

How to deal with `NA`'s? Many functions in R like `sum()`, `mean()`, and `median()` have the so-called `na.rm` argument to specify if missing values should be removed before any computation this feature. We can take advantage of `na.rm = TRUE`:

```{r}
# second round
standardize <- function(x) {
  x_mean <- mean(x, na.rm = TRUE)
  x_sd <- sd(x, na.rm = TRUE)
  x_centered <- x - x_mean
  z <- x_centered / x_sd
  return(z)
}

standardize(ht10)

standardize(starwars$height)
```

Now `standardize()` is able to return a more useful output by removing missing values. However, we should let the user decide if `NA`'s must be removed. We can include an argument in `standardize()` to indicate if missing values are to be removed:

```{r}
# third round
standardize <- function(x, na_rm = FALSE) {
  x_mean <- mean(x, na.rm = na_rm)
  x_sd <- sd(x, na.rm = na_rm)
  x_centered <- x - x_mean
  z <- x_centered / x_sd
  return(z)
}

# default call
standardize(starwars$height)

# removing NAs
standardize(starwars$height, na_rm = TRUE)
```


### Simplifying the body

So far we have a working function `standardize()` that does the job and takes care of potential missing values. We can take a further step and review the code of the body. Let's go back to the initial code:

```{r}
x <- ht10
x_mean <- mean(x)
x_sd <- sd(x)
x_centered <- x - x_mean
z <- x_centered / x_sd
```

The code above works, but it is very "verbose". We can take advantage of R's functional behavior to shorten the computation of the standard scores in one line:

```{r}
x <- ht10
z <- (x - mean(x)) / sd(x)
z
```

Having simplified the code, we can simplify our function:

```{r}
# fifth round
standardize <- function(x, na_rm = FALSE) {
  z <- (x - mean(x, na.rm = na_rm)) / sd(x, na.rm = na_rm)
  return(z)
}

standardize(tail(starwars$height, n = 10), na_rm = TRUE)
```

-----

# Documenting Functions

The examples of functions in this tutorial are simple, and fairly understandble (I hope so). However, you should strive to always include _documentation_ for your functions. What does this mean? Documenting a function involves adding descriptions for the purpose of the function, the inputs it accepts, and the output it produces. 

- Description: what the function does
- Input(s): what are the inputs or arguments
- Output: what is the output (returned value)

You can find some inspiration in the `help()` documentation when your search 
for a given function's description.

There are several approaches for writing documentation of a function. I will show you how to use what are called __roxygen comments__ to achieve this task. While not used by most useRs, they are great when you want to take your code and make a package out of it. 

Here's an example of documentation for `standardize()`

```{r}
#' @title Standardize
#' @description Transforms values in standard units (i.e. standard scores)
#' @param x numeric vector
#' @param na_rm whether to remove missing values
#' @return standardized values
#' @examples
#'   standardize(rnorm(10))
standardize <- function(x, na_rm = FALSE) {
  z <- (x - mean(x, na.rm = na_rm)) / sd(x, na.rm = na_rm)
  return(z)
}
```

- Roxygen comments are R comments formed by the hash symbol immediately followed by an apostrophe: `#'`

- You specify the label of a field with `@` and a keyword: e.g. `@title`

- The syntax highlighting of RStudio recognizes this type of comments and labels

- Typical roxygen fields:

| label          | meaning     | description                |
|----------------|-------------|----------------------------|
| `@title`       | title       | name of your function      |          
| `@description` | description | what the function does     |
| `@param input` | parameter   | describe `input` parameter |
| `@return`      | output      | what is the returned value | 

-----

### General Strategy for Writing Functions

- Always start simple with test toy-values.
- Get what will be the body of the function working first.
- Check out each step of the way.
- Don't try and do too much at once.
- Create (encapsulate body) the function once everything works.
- Optional: after you have a function that works, you may worry about "elegance", "efficiency", "cleverness", etc
- Include documentation; we suggest using Roxygen comments.
