---
title: "Programming Classes and Methods"
subtitle: "Stat 133, Spring 2018"
author: "Gaston Sanchez"
output: pdf_document
fontsize: 12pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

> This tutorial is the first part for the material to create an R package.
> In this document, we'll describe functions for tossing a coin. Specifically,
> we will define an object of class `"coin"`, and a function `"toss()"` for
> tossing a `"coin"`, and producing an object of class `"toss"`. We will also
> discuss how to create common methods such as `print()`, `summary()`, `plot()`,
> replacement, extraction, testing, and addition.


## 1) Tossing a Coin

To illustrate the concepts behind object-oriented programming in R, we are 
going to implement code that simulates tossing a coin one or more times. 


### 1.1) Coin object

To toss a coin using R, we first need an object that plays the role of a coin.
So let's start by creating a `coin` object using a character vector with two
elements: `"heads"` and `"tails"`:

```{r coin-vector}
# coin object
coin <- c("heads", "tails")
coin
```


### 1.2) Tossing a coin with `sample()`

Tossing a coin is a random experiment: you either get heads or tails. To get
a random output in R we can use the function `sample()` which takes a random
sample of a given vector. Here's how to simulate a coin toss using `sample()` 
to take a random sample of size 1 from `coin`:

```{r coin-sample}
# one toss
sample(coin, size = 1)
```

\bigskip

We can also use `sample()` to take samples of sizes different than one, and 
also to sample with replacement. To simulate multiple tosses, we can change the 
value of the `size` argument, and set `replace = TRUE`:

```{r three-tosses}
# 3 tosses
sample(coin, size = 3, replace = TRUE)
```

\bigskip

```{r six-tosses}
# 6 tosses
sample(coin, size = 6, replace = TRUE)
```


### 1.3) Function `toss()`

To make our code reusable, it's better to create a function that lets us 
toss a coin multiple times; for instance, we can write the following `toss()` 
function:

```{r toss-function1}
#' @title coin toss function 
#' @description simulates tossing a coin a given number of times
#' @param coin object (a vector)
#' @param times number of tosses
#' @return vector of tosses
toss <- function(coin, times = 1) {
  sample(coin, size = times, replace = TRUE)
}

# default call (1 toss)
toss(coin)

# 5 tosses
toss(coin, times = 4)
```

\bigskip

We can make the function more versatile by adding a `prob` argument that let us 
specify different probabilities for _heads_ and _tails_

```{r toss-function2}
#' @title coin toss function 
#' @description simulates tossing a coin a given number of times
#' @param coin object (a vector)
#' @param times number of tosses
#' @param prob vector of proabilities for each side of the coin
#' @return vector of tosses
toss <- function(coin, times = 1, prob = NULL) {
  sample(coin, size = times, replace = TRUE, prob = prob)
}

# toss a loaded coin 10 times
toss(coin, times = 10, prob = c(0.8, 0.2))
```


### 1.4) Computing Frequencies

Typical probability problems that have to do with coin tossing, require
to compute the total proportion of `"heads"` and `"tails"`:

```{r five-tosses}
# five tosses
five <- toss(coin, times = 5)

# proportion of heads and tails
sum(five == "heads") / 5
sum(five == "tails") / 5
```

\bigskip

It is also customary to compute the relative frequencies of `"heads"` and
`"tails"` in a series of tosses:

```{r relative-freqs}
# relative frequencies of heads
cumsum(five == "heads") / 1:length(five)

# relative frequencies of tails
cumsum(five == "tails") / 1:length(five)
```


## 1.5) Plotting relative frequencies

Likewise, it is common to look at how the relative frequencies of heads or 
tails change over a series of tosses:

```{r plot-freqs}
set.seed(5938)
hundreds <- toss(coin, times = 500)
head_freqs = cumsum(hundreds == "heads") / 1:500

plot(1:500, head_freqs, type = "l", ylim = c(0, 1), las = 1,
     col = "#3989f8", lwd = 2)
```

So far we have written code in R that simulates tossing a coin one or more
times. We have included commands to compute proportion of heads and tails, 
as well the relative frequencies of heads (or tails) in a series of tosses.
In addition, we have produced a plot of the relative frequencies and see
how, as the number of tosses increases, the frequency of heads (and tails) 
approach 0.5.

In the following sections, we are going to see how to implement various
functions and methods in R to make our coin tossing code more reliable,
more structured, and more useful.



## 2) Object-Oriented Programming

Popular languages that use OOP include C++, Java, and Python. Different 
languages implement OOP in different ways. R also provides OOP capabilities, 
but compared to other languages, R's OOP options are less formal.

The idea of OOP is that all operations are built around objects, which have 
a __class__, and __methods__ that operate on objects in the class. Classes are
constructed to build on (inherit from) each other, so that one class may be 
a specialized form of another class, extending the components and methods
of the simpler class (e.g. `"lm"`, and `"glm"` objects).

Often when you get to the point of developing OOP code in R, you're doing some
serious programming, and you're going to be acting as a software engineer.
It's a good idea to think carefully in advance about the design of the classes
and methods.


### 2.1) Programming without objects, classes, and methods

If you have no previous experience with object-oriented programming, it can 
be a bit challenging. You may be tempted to think that OOP does not provide
any evident advantages: you need to think in advance before writing code, 
brainstorm, choose the right objects, their types, what their relationships
will be, ... "so many things to consider". 
So why bother? Why care about objects?

Let me show you a simple example of why OOP is not a bad idea. 
Taking the code we've written for tossing a coin, we can generate two series 
of tosses. The first experiment involves tossing a coin five times, and then 
computing the proportion of heads:

```{r}
# random seed
set.seed(534)

# five tosses
five <- toss(coin, times = 5)

# prop of heads in five
sum(five == "heads") / length(five)
```

\bigskip

The second experiment involves tossing a coin six times and computing the 
proportion of heads:

```{r}
# six tosses
six <- toss(coin, times = 6)

# prop of heads in six
sum(six == "heads") / length(five)
```

\bigskip

The above code works ... except that there is an error; the number of heads in
`six` is being divided by 5 instead of 6. R hasn't detected this error: it
doesn't know that the division has to be done using `length(six)`.

Wouldn't it be prefarable to have some mechanism that prevented this type
of errors from happening? Bugs will always be part of any programming 
activity, but it is better to minimize certain types of errors like the one
above.


### 2.2) Object-Oriented Programming in R

R has two (plus one) object oriented systems, so it can be a bit intimidating
when you read and learn about them for the first time. 
The goal of this tutorial is not to make you an expert in all R's OOP systems, 
but to help you become familiar with the so-called _S3 class_.

__S3__ implements a style of OO programming called generic-function OO. 
S3 uses a special type of function called a _generic_ function that decides 
which method to call, e.g., `table(iris$Species)`. Keep in mind that S3 is a 
very casual system: it does not really have a formal definition of classes.

S3 classes are widely-used, in particular for statistical models in the 
`"stats"` package. S3 classes are very informal in that there is not a formal
definition for an S3 class. Usually, S3 objects are built on top of lists, or 
atomic vectors with attributes. But you can also turn functions into S3 objects.

Note that in more formal OOP languages, all functions are associated with a 
class, while in R, only some are.


### 2.3) S3 Classes and Objects

To make an object an instance of a class, you just take an existing base object 
and set the `"class"` attribute. You can do that during creation of the object
with `structure()`, or after the object has been createad with `class()`.

```{r structure-class}
# object coin via structure()
coin1 <- structure(c("heads", "tails"), class = "coin") 

# object coin via class()
coin2 <- c("heads", "tails")
class(coin2) <- "coin" 
```

You can inspect the class of any object using `class(x)`

```{r class-example}
class(coin1)
```

You can also determine if an object inherits from a specific class using
`inherits()`

```{r inherits-example}
inherits(coin2, "coin")
```


### 2.4) Generic and Specific Methods

Having a coin object, we can simulate flipping the coin with our previously 
defined function `toss()`.

```{r}
toss(coin1, times = 5)
```

The issue with the way `toss()` is defined, is that you can pass it any type
of vector (not necessarily of class `"coin"`), and it will still work:

```{r}
toss(c('tic', 'tac', 'toe'))
```

\bigskip

To create a function `toss()` that only works for objects of class `"coin"`, we
could add a `stop()` condition that checks if the argument `coin` is of the
right class:

```{r}
toss <- function(coin, times = 1, prob = NULL) {
  if (class(coin) != "coin") {
    stop("\nflip() requires an object 'coin'")
  }
  sample(coin, size = times, replace = TRUE, prob = prob)
}

# ok
toss(coin1)

# bad coin
toss(c('tic', 'tac', 'toe'))
```

\bigskip

A more formal strategy, and one that follows OOP principles, is to create a
toss __method__. In R, many functions are actually methods: e.g. `print()`,
`summary()`, `plot()`, `str()`, etc. 

```{r print-method}
# print method
print
```

\bigskip

```{r plot-method}
# plot method
plot
```

These types of functions are not really one unique function, they typically 
comprise a collection or family of functions for printing objects, computing 
summaries, plotting, etc. Depending on the class of the object, a generic 
method will look for a specific function for that class:

```{r matrix-methods}
# methods for objects "matrix"
methods(class = "matrix")
```


### 2.5) `toss` method

When implementing new methods, you begin by creating a __generic__ method with 
the function `UseMethod()`:

```{r toss-method}
toss <- function(x, ...) UseMethod("toss")
```

The function `UseMethod()` allows you to declare the name of a method. In this
example we are telling R that the function `toss()` is now a generic `"toss"` 
method. Note the use of `"..."` in the function definition, this will allow you 
to include more arguments when you define specific methods based on `"toss"`. 

A generic method alone is not very useful. You need to create specific
cases for the generic. In our example, we only have one class `"coin"`,
so that is the only class we will allow `toss` to be applied on. The way to
do this is by defining `toss.coin()`:

```{r toss-coin-specific}
toss.coin <- function(coin, times = 1, prob = NULL) {
  sample(coin, size = times, replace = TRUE, prob = prob)
}
```

The name of the method, `"toss"`, comes first, followed by a dot `"."`, 
followed by the name of the class, `"coin"`. Notice that the body of the 
function `toss.coin()` does not include the `stop()` command anymore.

To use the `toss()` method on a `"coin"` object, you don't really have to call 
`toss.coin()`; calling `toss()` is enough:

```{r}
toss(coin1)
```

How does `toss()` work? Becasue `toss()` is now a generic method, everytime
you use it, R will look at the class of the input, and see if there is an
associated `"toss"` method. In the previous example, `coin1` is an object 
of class `"coin"`, for which there is a specific `toss.coin()` method. 
Thus using `toss()` on a `"coin"` object works fine. 

Now let's try `toss()` on the character vector `c('tic', 'tac', 'toe')`:

```{r}
# no toss() method for regular vectors
toss(c('tic', 'tac', 'toe'))
```

When you try to use `toss()` on an object that is not of class `"coin"`, 
you get a nice error message.



## 3) A more robust `"coin"` class

Let's review our class `"coin"`. The way we defined a coin object was like this:

```{r simple-coin}
# object coin
coin1 <- c("heads", "tails")
class(coin1) <- "coin" 
```

While this definition is good to illustrate the concept of an object, its 
class, and how to define generic methods, it is a very loose-defined class.
One could create a `"coin"` out of `c('tic', 'tac', 'toe')`, and then use 
`toss()` on it:

```{r}
ttt <- c('tic', 'tac', 'toe')
class(ttt) <- "coin"

toss(ttt)
```

We need a more formal definition of a coin. For instance, it
makes more sense to require that a coin should only have two sides. In this way, 
the vector `ttt` would not be a valid coin. 

For convenience purposes, we can define a __class constructor__ function to
initialize a `"coin"` object:

```{r}
coin <- function(object = c("heads", "tails")) {
  class(object) <- "coin"
  object
}

# default coin
coin()

# another coin
coin(c("h", "t"))
```


### 3.1) Improving `"coin"` objects

To implement the requirement that a coin must have two sides, we can check 
for the length of the input vector:

```{r}
coin <- function(object = c("heads", "tails")) {
  if (length(object) != 2) {
    stop("\n'object' must be of length 2")
  }
  class(object) <- "coin"
  object
}

# US penny
penny <- coin(c("lincoln", "shield"))
penny
```

\bigskip

```{r invalid-coin}
# invalid coin
coin(ttt)
```

\bigskip

Because the `toss()` function simulates flips using `sample()`, we can take 
advantage of the argument `prob` to specify probabilities for each side of 
the coin. In this way, we can create _loaded_ coins. 

We can add a `prob` argument to the constructor function. This argument takes a 
vector of probabilities for each element in `object`, and we pass this vector
as an attribute of the coin object. Furthermore, we can set a default 
`prob = c(0.5, 0.5)`, that is, a _fair_ coin by default:

```{r}
coin <- function(object = c("heads", "tails"), prob = c(0.5, 0.5)) {
  if (length(object) != 2) {
    stop("\n'object' must be of length 2")
  }
  attr(object, "prob") <- prob
  class(object) <- "coin"
  object
}

coin()
```

\bigskip

Once again, we need to check for the validity of `prob`. Here is one possible
function to check several aspects around `prob`: must be of numeric and of 
length 2, probability values must be between 0 and 1, and the sum of these 
values must add up to 1:

```{r check-prob}
check_prob <- function(prob) {
  if (length(prob) != 2 | !is.numeric(prob)) {
    stop("\n'prob' must be a numeric vector of length 2")
  }
  if (any(prob < 0) | any(prob > 1)) {
    stop("\n'prob' values must be between 0 and 1")
  }
  if (sum(prob) != 1) {
    stop("\nelements in 'prob' must add up to 1")
  }
  TRUE
}
```

Note that I'm adding a `TRUE` statement at the end of the function. This is
just an auxiliary value to know if the function returns a valid `prob`.
Now let's test it:

```{r check-good-probs, results='hide'}
# good prob
check_prob(c(0.5, 0.5))
check_prob(c(0.1, 0.9))
check_prob(c(1/3, 2/3))
check_prob(c(1/3, 6/9))
```

```{r check-bad-probs}
# bad length
check_prob(1)
# bad length
check_prob(c(0.1, 0.2, 0.3))
# negative probability
check_prob(c(-0.2, 0.8))
# what should we do in this case?
check_prob(c(0.33, 0.66))     
```

\bigskip

Here's the improved constructor function `coin()`:

```{r}
coin <- function(object = c("heads", "tails"), prob = c(0.5, 0.5)) {
  if (length(object) != 2) {
    stop("\n'object' must be of length 2")
  }
  check_prob(prob)
  attr(object, "prob") <- prob
  class(object) <- "coin"
  object
}

coin1 <- coin()
coin1
```

\bigskip

And the new definition of `toss.coin()`:

```{r new-toss}
toss.coin <- function(coin, times = 1) {
  sample(coin, size = times, replace = TRUE, prob = attr(coin, 'prob'))
}
```

Let's toss a loaded coin:

```{r}
set.seed(2341)
load_coin <- coin(c('HEADS', 'tails'), prob = c(0.75, 0.25))
toss(load_coin, times = 6)
```


### 3.2) Extending classes

We can extend the class `"coin"` and create a derived class for special types
of coins. For instance, say we want to create a class `"quarter"`. One side
of the coin refers to George Washington, while the other side refers to John
Brown's Fort:

https://en.wikipedia.org/wiki/Quarter_(United_States_coin)

```{r quarter-coin}
quarter1 <- coin(c("washington", "fort")) 
class(quarter1) <- c("quarter", "coin")
quarter1
```

Our coin `quarter1` inherits from `"coin"`:

```{r}
inherits(quarter1, "coin")
```

Likewise, we can create a class for a slightly unbalanced `"dime"`:

```{r dime-coin}
dime1 <- coin(c("roosevelt", "torch"), prob = c(0.48, 0.52))
class(dime1) <- c("dime", "coin")
dime1
```



## 4) Object `"toss"`

Because we are not only interested in tossing a coin, but also in keeping track
of such tosses, it would be good to have another object for this purpose.
How do you know that you need this new object class? Well, this is precisely
an example that illustrates the process of programming in general, and OOP in
particular. This kind of decisions require some (or a lot of) thinking, 
and brainstorming time. The more you understand a problem (i.e. phenomenon,
process), the better you will be prepared to design what objects and methods
you need to program.

While I was writing this material, I decided that it would be convenient to
have an object of class `"toss"` containing the following information:

- all the outcomes from the series of tosses
- the total number of tosses
- the total number of heads
- the total number of tails

The most flexible type of data structure in R to store other data structures 
is a `list`. Having a vector of tosses, we can use a list to keep all the
desired information:

```{r}
flips <- toss(coin1, times = 6)

a <- list(
  tosses = flips, 
  total = length(flips),
  heads = sum(flips == "heads"),
  tails = sum(flips == "tails")
)

a
```

For convenience purposes, we can write an __auxiliary constructor__ function, 
which I will call `make_toss()`. This function takes an input vector (i.e. a
character vector with `"heads"` and `"tails"` elements), and it returns an
object of class `"toss"`:

```{r}
make_toss <- function(coin, flips) {
  res <- list(
    coin = coin,
    tosses = flips,
    total = length(flips),
    heads = sum(flips == coin[1]),
    tails = sum(flips == coin[2]))
  class(res) <- "toss"
  res
}
```


### 4.1) Main Function `toss()`

Now that we have the auxiliary function `make_toss()`, we can encapsulate it in 
a _master_ function `toss.coin()`:

```{r toss-ver1}
toss.coin <- function(coin, times = 1) {
  flips <- sample(coin, size = times, replace = TRUE, prob = attr(coin, 'prob'))
  make_toss(coin, flips)
}
```

```{r}
set.seed(2233)
fair <- coin()
toss(fair, times = 5)
```

\bigskip

You may ask why we need a function `make_toss()`, and another function `toss()`. 
Can't we just write a single function `suppertoss()` that does everything at once?:

```{r}
supertoss <- function(coin, times = 1) {
  flips <- flip(coin, times = times)
  res <- list(
    coin = coin,
    tosses = flips,
    total = length(flips),
    heads = sum(flips == coin[1]),
    tails = sum(flips == coin[2]))
  class(res) <- "toss"
  res
}
```

The short answer is: yes, you can. And probably this is what most beginners
tend to do. The reason why I decided to break things down into simpler and
smaller functions is because I went already through a couple of implementations,
and realized that it was better to have the auxiliary function `make_toss()`. 
Also, it is good practice to write short functions that preferably do 
one thing. 

Here's a brief recap of the main functions we have so far:

- `coin()` is a constructor function to create objects of class `"coin"`.
- `toss()` is a generic `"toss"` method.
- `make_toss()` is an auxiliary function that takes a `"coin"` and a vector of 
flips, and which produces an object `"toss"`.
- `toss.coin()` is the specific `"toss"` method to be used on `"coin"` objects.


## 5) Upgrading `toss()`

Let's consider our `quarter` coin, and apply `toss()` on it:

```{r quarter}
quarter1 <- coin(c("washington", "fort")) 
class(quarter1) <- c("quarter", "coin")
quarter1

toss(quarter1, times = 4)
```

\bigskip

`toss()` is working as expected, and you can try it with different values for 
`times`. The only issue is that a distracted user could pass an unexpected value 
for the argument `times`:

```{r}
toss(quarter1, times = -4)
```

R produces an error when `times = -4`, but it's an error that may not be very
helpful for the user. The error message clearly says that `'size'` is an 
invalid argument, but `toss()` just has one argument: `times`.

To be more user friendly, among other reasons, it would be better to check 
whether `times` has a valid value. One way to do that is to include a 
conditional statement:

```{r toss-ver2}
toss.coin <- function(coin, times = 1) {
  if (times <= 0) {
    stop("\nargument 'times' must be a positive integer")
  }
  flips <- sample(coin, size = times, replace = TRUE, prob = attr(coin, 'prob'))
  make_toss(coin, flips)
}

# this works ok
toss(quarter1, 5)

# this doesn't work, but the error message is clear
toss(quarter1, -4)
```

Once again, it is good practice to write short functions that preferably do 
one thing. In this case, we could define a checking function for `times`:

```{r}
# auxiliary function to check 'times' input
check_times <- function(times) {
  if (times <= 0 | !is.numeric(times)) {
    stop("\nargument 'times' must be a positive integer")
  } else {
    TRUE
  }
}
```

and then include `check_times()` inside `toss()`:

```{r toss-ver3}
toss.coin <- function(coin, times = 1) {
  check_times(times)
  flips <- sample(coin, size = times, replace = TRUE, prob = attr(coin, 'prob'))
  make_toss(coin, flips)
}

toss(quarter1, 5)
```



## 6) More Methods

Let's see how to create common R methods for a `"toss"` object such as
`print()`, `'summary()`, and `plot()`.


### 6.1) Print Method

Typically, most classes in R have a dedicated printing method. To create such
a method we use the generic function `print()` like so:

```{r print-toss}
# print method for object of class "toss"
print.toss <- function(x, ...) {
  cat('object "toss"\n')
  cat(sprintf('coin: "%s", "%s"', x$coin[1], x$coin[2]), "\n")
  cat("total tosses:", x$total, "\n")
  cat(sprintf("num of %s:", x$coin[1]), x$heads, "\n")
  cat(sprintf("num of %s:", x$coin[2]), x$tails, "\n")
  invisible(x)
}
```

By convention, `print` methods return the value of their principal argument
invisibly. The `invisible` function turns off automatic printing, thus 
preventing an infinite recursion when printing is done implicitly at the 
session level.

After a `print` method has been defined for an object `"toss"`, everytime
you type an object of such class, R will search for the corresponding method
and display the output accordingly:

```{r}
# testing print method
quarter_flips <- toss(quarter1, 50)
quarter_flips
```


### 6.2) Summary Method

For most purposes the standard `print` method will be sufficient output, but
sometimes a more extensive display is required. This can be done with a
`summary`. To define this type of method we use the function `summary()`.

```{r summary-toss}
summary.toss <- function(object) {
  structure(object, class = c("summary.toss", class(object)))
}

print.summary.toss <- function(x, ...) {
  cat('summary "toss"\n\n')
  cat(sprintf('coin: "%s", "%s"', x$coin[1], x$coin[2]), "\n")
  cat("total tosses:", x$total, "\n\n")
  cat(sprintf("num of %s:", x$coin[1]), x$heads, "\n")
  cat(sprintf("prop of %s:", x$coin[1]), x$heads/x$total, "\n\n")
  cat(sprintf("num of %s:", x$coin[2]), x$tails, "\n")
  cat(sprintf("prop of %s:", x$coin[2]), x$tails/x$total, "\n")
  invisible(x)
}
```

Let's test it:

```{r}
summary(quarter_flips)
```


### 6.3) Plot Method

We can also define a `plot` method for objects of class `"toss"`:

What we want to plot of an object `"toss"` is the series of realtive frequencies 
(of either `"heads"` ot `"tails"`). This means we need to create a couple of
auxiliary functions:

```{r}
head_freqs <- function(x) {
  cumsum(x$tosses == x$coin[1]) / 1:x$total
}


tail_freqs <- function(x) {
  cumsum(x$tosses == x$coin[2]) / 1:x$total
}


frequencies <- function(x, side = 1) {
  if (side == 1) {
    return(head_freqs(x))
  } else {
    return(tail_freqs(x))
  }
}
```

Here's one way to define a `plot()` method for `"toss"` objects:

```{r plot-toss}
plot.toss <- function(x, side = 1, ...) {
  freqs <- frequencies(x, side = side)
  plot(1:x$total, freqs, type = "n", ylim = c(0, 1), las = 1,
       xlab = "number of tosses", bty = "n",
       ylab = sprintf("relative frequency of %s", x$coin[side]))
  abline(h = 0.5, col = "gray70", lwd = 1.5)
  lines(1:x$total, freqs, col = "tomato", lwd = 2)
  title(sprintf("Relative Frequencies in a series of %s coin tosses", x$total))
}
```

Let's test our `plot` method:

```{r}
plot(quarter_flips)
```



## 7) Additional Methods

Until now we have standard methods like `print()`, `summary()`, and `plot()`.
However, we can add more common (and not so common) methods such as:

- replacement: `"[<-.toss"`
- extraction: `"[.toss"`
- testing: `is.toss()`
- addition: `"+.toss"`


### 7.1) Replacement Method

Replacement functions are those calls like `x[1] <- 3`. The function behind 
this expression is the replacement `"[<-"()` function.
We can also create a replacement function for a given class using the notation
`"[<-.class"`, where `class` is the name of the class:

```{r replacement-toss}
"[<-.toss" <- function(x, i, value) {
  if (value != x$coin[1] & value != x$coin[2]) {
    stop(sprintf('\nreplacing value must be %s or %s', x$coin[1], x$coin[2]))
  }
  x$tosses[i] <- value
  make_toss(x$coin, x$tosses)
}
```

Test it:

```{r test-replacement}
set.seed(3752)
b <- toss(dime1, times = 5)
b$tosses

# replacement
b[1] <- "torch"
b$tosses
```

What about replacing out of the original range?

```{r weird-test-replacement}
# replacement in sixth position (weird!)
b[6] <- "torch"
b
```

Or something like this?

```{r fail-test-replacement}
# replacement
b[10] <- "torch"
b
```

\bigskip

Because it does not make sense to replace if index is out of the original length,
we can add a `stop()` condition:

```{r better-replacement-toss}
"[<-.toss" <- function(x, i, value) {
  if (value != x$coin[1] & value != x$coin[2]) {
    stop(sprintf('\nreplacing value must be %s or %s', x$coin[1], x$coin[2]))
  }
  if (i > x$total) {
    stop("\nindex out of bounds")
  }
  x$tosses[i] <- value
  make_toss(x$coin, x$tosses)
}
```

Now we cannot replace if index is out of the original length:

```{r error-replacement}
set.seed(3752)
b <- toss(dime1, times = 5)
b$tosses

# replacement
b[10] <- "torch"
```


### 7.2) Extraction Method

What if you want to know what is the value of toss in position 3? You could
type something like this:

```{r}
b$tosses[3]
```

Or you could create an extraction method that allows you to type `x[3]`. 
The function behind this expression is the extraction `"["()` function.
We can also create a extraction function for a given class.

```{r extract-toss}
"[.toss" <- function(x, i) {
  x$tosses[i]
}
```

Test it:

```{r test-extract}
set.seed(3752)
b <- toss(dime1, times = 5)
b$tosses
b[1]
```


### 7.3) Is `"toss"`

Another common type of function for an object of a given class is 
`is.class()`-like functions: e.g. `is.list()`, `is.numeric()`, `is.matrix()`.

```{r}
is.toss <- function(x) {
  inherits(x, "toss")
}

is.toss(b)
is.toss(c("heads", "tails"))
```



### 7.4) Addition Method

R comes with generic Math methods (see `?Math`). Among these generic methods 
we can find the `"+"` operator. This means that we can define our own _plus_
method for objects of class `"toss"`. The idea is to be able to call a command
like this:

```{r eval = FALSE}
# toss object
b <- toss(dime1, times = 5)

# add 5 more flips
b + 5
```

Here's one implementation of `"+.toss()"` in which the first argument is an 
object of class `"toss"`, and the second argument is a single positive number
that will play the role of additional tosses:

```{r}
"+.toss" <- function(obj, incr) {
  if (length(incr) != 1 | incr <= 0) {
    stop("\ninvalid increament (must be positive)")
  }
  more_flips <- toss(obj$coin, times = incr)
  make_toss(obj$coin, c(obj$tosses, more_flips$tosses))
}
```

Remember that `"+"` is a binary operator, which means that writing a `"+"` 
method requires a function with two arguments. Let's try it:

```{r}
# add four more tosses
mycoin <- coin()
seven <- toss(mycoin, times = 7)
seven
```

\bigskip

Let's add a couple of more tosses to `seven`:

```{r}
# two more flips
seven + 2

# three more flips
seven + 3
```
